Semantic Linearity in a Lazy Optimising Compiler

### 2025-01-24 icfp_submission_2024.tex
added linearity-aware semantics

\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Gamma : \Lambda p.e \Downarrow \Gamma : \Lambda p.e}
    \qquad
    \infer[]
    {\Gamma : e\Downarrow \Delta : \Lambda p.e' \quad \Delta :
    e'[\pi/p]\Downarrow \Theta : v}
    {\Gamma : e~\pi\Downarrow \Theta : v }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma : \lambda x:_\pi \sigma . e \Downarrow \Gamma : \lambda x:_\pi \sigma . e }
    \qquad
    \infer[]
    {\Gamma : e \Downarrow \Delta : \lambda x:_\pi \sigma . e' \quad
    \Delta : e'[x/y] \Downarrow \Theta : v }
    {\Gamma : e~x \Downarrow \Theta : v}\\[1.5em]
    \infer[]
    {\Gamma : e \Downarrow \Delta : v}
    {(\Gamma , x:_\omega \sigma = e) : x \Downarrow (\Delta,x:_\omega
    \sigma = v) : v}
    \qquad
    \infer[]
    {(\Gamma,x:_\omega \sigma = e) : e' \Downarrow \Delta : v}
    {\Gamma : \llet{x{:}_{\Delta}\sigma = e}{e'}\Downarrow \Delta : v
    }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma : K~x_1~\dots~x_n \Downarrow \Gamma : K~x_1~\dots~x_n}
    \qquad
    \infer[]
    {\Gamma : e \Downarrow \Delta : K~x_1~\dots~x_n \quad
    \Delta : e'[x_i/y_i] \Downarrow \Theta : v }
    {\Gamma : \ccase{e}{z{:}_{\Delta}\sigma~\{\overline{\dots,K~y_1~\dots~y_n\to e'}\}}
    \Downarrow \Theta : v}\\[1.5em]
    \infer[]
    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % \Downarrow \Theta_n : v_n \quad
    (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e' \Downarrow \Theta
    : v
    }
    {\Gamma : \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'}
    \Downarrow \Theta : v}
    \end{array}
\]

We follow an approach
similar to that of \cite{cite:linearhaskell}.

>>>>

\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Gamma : \Lambda p.e \Downarrow \Gamma : \Lambda p.e}
    \qquad
    \infer[]
    {\Gamma : e\Downarrow \Delta : \Lambda p.e' \quad \Delta :
    e'[\pi/p]\Downarrow \Theta : v}
    {\Gamma : e~\pi\Downarrow \Theta : v }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma : \lambda x:_\pi \sigma . e \Downarrow \Gamma : \lambda x:_\pi \sigma . e }
    \qquad
    \infer[]
    {\Gamma : e \Downarrow \Delta : \lambda y:_\pi \sigma . e' \quad
    \Delta : e'[x/y] \Downarrow \Theta : v }
    {\Gamma : e~x \Downarrow \Theta : v}\\[1.5em]
    \infer[]
    {\Gamma : e \Downarrow \Delta : v}
    {(\Gamma , x:_\omega \sigma = e) : x \Downarrow (\Delta,x:_\omega
    \sigma = v) : v}
    \qquad
    \infer[]
    {(\Gamma,x:_\omega \sigma = e) : e' \Downarrow \Theta : v}
    {\Gamma : \llet{x{:}_{\Delta}\sigma = e}{e'}\Downarrow \Theta : v
    }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma : K~x_1~\dots~x_n \Downarrow \Gamma : K~x_1~\dots~x_n}
    \qquad
    \infer[]
    {\Gamma : e \Downarrow \Delta : K~x_1~\dots~x_n \quad
    \Delta : e'[x_i/y_i][K~x_1~\dots~x_n/z] \Downarrow \Theta : v }
    {\Gamma :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i}, {\dots} \to e'}\}}
    \Downarrow \Theta : v}\\[1.5em]
    \infer[]
    {\Gamma : e \Downarrow \Delta : K~x_1~\dots~x_n \quad
    \Delta : e'[K~x_1~\dots~x_n/z] \Downarrow \Theta : v }
    {\Gamma :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
    e'}\} \Downarrow \Theta : v}
    \qquad
    \infer[]
    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % \Downarrow \Theta_n : v_n \quad
    (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e' \Downarrow \Theta
    : v
    }
    {\Gamma : \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'}
    \Downarrow \Theta : v}
    \end{array}
\]

We follow an approach
similar to that of \cite{cite:linearhaskell}, defining a linearity
aware operational semantics in which linear variables are erased from
the runtime environment, ensuring that any term violating linearity
will result in a stuck state. The semantics is type-aware, defined by
the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
\mid v) : \sigma$ where $\Gamma$ and $\Delta$ are typing judgments
($\Gamma$ tracks unrestricted variables, $\Delta$ tracks linear
assumptions);
$\Theta$ and $\Theta'$ are runtime environments, consisting of bindings of the form
$x :_\pi \sigma = e$; $e$ is the expression being evaluated and $v$
its resulting value, both of type $\sigma$.

\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Gamma;\Delta \vdash (\Theta \mid \Lambda p.e) \Downarrow (\Theta \mid
    \Lambda p.e) : \forall p . \sigma }
    \\[1.5em]
    \infer[]
    {\Gamma ; \Delta \vdash (\Theta \mid  e) \Downarrow (\Theta' \mid
    \Lambda p.e') : \forall p . \sigma \quad \Gamma ;\Delta \vdash
    (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' : v) : \sigma}
    {\Gamma ; \Delta \vdash (\Theta \mid e~\pi) \Downarrow (\Theta''
    \mid v) : \sigma  }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma ;\Delta \vdash (\Theta \mid \lambda x:_\pi \sigma . e)
    \Downarrow (\Theta \mid \lambda x:_\pi\sigma .e) : \sigma
    \rightarrow_\pi \tau }\\[1.5em]
    \infer[]
    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
    \lambda y:_\pi \sigma . e') : \sigma \rightarrow_\pi \tau
    \quad
    \Gamma ; \Delta \vdash (\Theta' \mid e'[x/y]) \Downarrow (\Theta''
    : v) : \tau }
    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta'
    \mid v) : \tau}\\[1.5em]

       \infer[]
    {\Gamma , x:_{\Delta'} \sigma ; \Delta  \vdash (\Theta \mid e)
    \Downarrow (\Theta' \mid v) : \sigma}
    {\Gamma ; \Delta \vdash (\Theta , x:_{\Delta'} \sigma = e \mid x)
    \Downarrow (\Theta' , x:_{\Delta'} \sigma = v \mid v) : \sigma}
    \\[1.5em]
           \infer[]
    {\Gamma ;\Delta  \vdash (\Theta \mid e)
    \Downarrow (\Theta' \mid v) : \sigma}
    {\Gamma ; \Delta \vdash (\Theta , x:_1 \sigma = e \mid x)
    \Downarrow (\Theta' \mid v) : \sigma}
    \\[1.5em]

    \infer[]
    {\Gamma ;\Delta \vdash (\Theta,x:_{\Delta'} \sigma = e \mid e')
    \Downarrow (\Theta' \mid v) : \tau}
    {\Gamma ; \Delta \vdash (\Theta \mid \llet{x{:}_{\Delta}\sigma =
    e}{e'})\Downarrow (\Theta' \mid v) : \tau
    }\\[1.5em]

    \infer[]
    {\,}
    {\Gamma ; \Delta \vdash (\Theta \mid K~x_1~\dots~x_n) \Downarrow
    (\Theta \mid K~x_1~\dots~x_n) : T}
    \\[1.5em]

    \infer[]
    {\Gamma;\Delta \vdash (\Theta \mid e) \Downarrow (\Theta' :
    K~x_1~\dots~x_n) : T \quad
    \Gamma ;\Delta \vdash (\Theta' \mid e'[x_i/y_i][K~x_1~\dots~x_n/z]
    ) \Downarrow (\Theta'' : v) :\tau }
    {\Gamma ;\Delta \vdash (\Theta \mid
    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i}, {\dots} \to e'}\}})
    \Downarrow (\Theta'' \mid  v) : \tau}\\[1.5em]

    \infer[]
    {\Gamma ;\Delta \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
    K~x_1~\dots~x_n) : T \quad
    \Gamma ;\Delta \vdash (\Theta' \mid  e'[K~x_1~\dots~x_n/z])
    \Downarrow (\Theta'' \mid v) : \tau }
    {\Gamma ; \Delta \vdash (\Theta \mid 
    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
    e'}\}) \Downarrow (\Theta'' \mid v) : \tau}
    \\[1.5em]
    \infer[]
    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % \Downarrow \Theta_n : v_n \quad
    (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_\omega \sigma_i =
    e_i} \mid e' )\Downarrow (\Theta'
    \mid v) : \tau
    }
    {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'})
    \Downarrow (\Theta' \mid v) : \tau}

   \end{array}
\]
